Evaluation of 3-point correlation functions from structural images on CPU and GPU architectures: accounting for anisotropy effects

### 2024-08-29 paper.tex
web-like -> weblike + fill-region

\subsection{Triangular sampling pattern}
There are three major known sampling pattern implementations reported in the literature (\cref{fig:pattern}):
\begin{enumerate}
  \item Arbitrary pattern (or random) triangle is thrown into the structure (e.g., \cite{malmir2018});
  \item Web-like predefined triangular pattern (e.g., \cite{SMITH1988176});
  \item Right-triangular sampling pattern (e.g., \cite{dimitrakopoulos2010high}).
\end{enumerate}
The second case of the web-like pattern was useful at times of low computational resources and seems to be the least flexible in terms of structure coverage. Right-traingle as a pattern was popular due ease of implementation and potential speed gains due to vectorization.
However, we have organized the algorithms of \code{CorrelationFunctions.jl} to be able to utilize any patterns.
This was achieved by organizing computational functions not around a shifting pattern, but rather around precomputed points representing
triangle's vertices -- thus making our approach under the hood using the ideas behind the web-like pattern method.

@@

\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.8\linewidth]{images/web-like.png}
  \caption[]{Schematic representation of patterns for 3-point correlation functions evaluation:
  arbitrary or random triangle, web-like, and right-triangular patterns. Note that real patterns are
  more complex and are made simple for visualization, i.e., for right-triangular pattern we sample
  all possible variations between points. Equilateral or isosceles triangles can be considered
  as arbitrary patterns and can be either fixed in space, or rotated randomly.}
  \label{fig:pattern}
\end{figure*}

@@

\subsection{Comparison of sampling patterns}
\label{sec:compare_patterns}
To compare different sampling and averaging patterns we performed computations of $S_3$ on Sandstone 3D sample.
They were performed in three different manners: 1) isosceles triangle with a right-triangle
between the equal side -- this sampling resembles both our right-triangle pattern and pattern
of Malmir et al. \cite{malmir2018}, computations were made within XY planes within the image and, thus,
represent directional CF; 2) equilateral triangle with side $r$ directionally analyzed within XY planes;
3) equilateral triangle with side $r$ averaged over 20 random rotations -- reminiscence of web-like and random patterns.

@@

\subsection{Applications to material images}

\begin{table*}[!htp]
  \centering
  \begin{tabular}{|c|c|c|c|c|}
    \hline
    Sample name & Image type & Dimensions (pixels/voxels) & Image resolution ($\mu m$)
    & $Q_{0.5}$ \\
    \hline
    Sandstone & XCT &  $500 \times 500 \times 500$ & 2.25 & 0.9703 \\
    Soil & Reconstruction & $500 \times 500 \times 500$ & 0.72 & 0.9431 \\
    Shale & Reconstruction & $160 \times 160 \times 160$ & 2.25 & 0.8912 \\
    Sandstone & SEM &  $1280 \times 869$ & 0.10 & 0.9613 \\
    \hline
  \end{tabular}
  \caption{Summary of samples used for calculation of correlation functions.}
  \label{tab:summary}
\end{table*}
In this section we compute correlation functions for 2- and 3-dimensional porous
media samples. Table \ref{tab:summary} contains information about these
samples. Note that we compute surface correlation functions only for those
samples for which the criterion $Q_{0.5} > 0.97$ holds (the two sandstones and
the soil) \cite{samarin2023robust}. Images of the samples are presented on
\cref{fig:samples}. On plots of correlation functions for isotropic samples we
provide a family of curves with the first argument of the correlation function
being fixed. For anisotropic samples we provide heatmaps of correlation
functions where anisotropy can be easily observed.
\begin{figure*}[tp]
  \centering
  \subfigure[Sanstone XCT]{
    \includegraphics[width=0.4\linewidth]{images/15581.png}
    \label{fig:sample-sandstone-xct}}
  \hfill
  \subfigure[Sandstone SEM (different clusters of void phase are shown in color)]{
    \includegraphics[width=0.4\linewidth, frame]{images/sandstone1.png}
    \label{fig:sample-sandstone}}
  \vskip\baselineskip
  \subfigure[Soil]{
    \includegraphics[width=0.4\linewidth]{images/soil.png}
    \label{fig:sample-soil}}
  \hfill
  \subfigure[Shale]{
    \includegraphics[width=0.4\linewidth]{images/shale.png}
    \label{fig:sample-shale}}
  \caption[]{Images of samples used for calculation of higher-order correlation
    functions.}
  \label{fig:samples}
\end{figure*}

@@

We have made execution time comparisons of our algorithm with existing
algorithms.  However, it is also beneficial to compare some algorithmic details
against existing digital approaches -- proposed by Malmir~et.~al.
\cite{malmir2018} and another one by Smith and Torquato \cite{SMITH1988176}. The
first algorithm generates a triangle with fixed sides $r_1$ and $r_2$ and an
arbitrary angle $\theta$ between them and places it in the medium firstly making
a random translation and a rotation around a random vector by a random angle. By
making a large number of trials the algorithm gets a value of a three-point
correlation function at the point $(r_1, r_2, \theta)$. The second algorithm
takes a precalculated triangle $(r_1, r_2, \theta)$ from a web-like pattern and
places it in the medium without rotation, only applying a random
translation. This way the algorithm preserves orientation of triangles stored in
the pattern. The main advantage of our approach that can operate on arbitrary patterns is in
its parallelizability and ease of implementation. The rotations necessary
for arbitrary placement of the pattern within the image can be random as in the Malmir et al. algorithm, or it
can be a single rotation to align the sampling triangle with the
direction of anisotropy (\cref{fig:aniso}). The latter allows computing
directional 3-point correlation functions for anisotropic structures similar to
2-point counterparts \cite{10.1063/1.4867611,EPL1}.

>>>>

\subsection{Triangular sampling pattern}
There are three major known sampling pattern implementations reported in the
literature (\cref{fig:pattern}):
\begin{enumerate}
  \item Arbitrary pattern (or random) triangle is thrown into the structure
    (e.g., \cite{malmir2018});
  \item Weblike predefined triangular pattern (e.g., \cite{SMITH1988176});
  \item Right-triangular sampling pattern (e.g.,
    \cite{dimitrakopoulos2010high}).
\end{enumerate}
The second case of the weblike pattern was useful at times of low computational
resources and seems to be the least flexible in terms of structure
coverage. Right-traingle as a pattern was popular due ease of implementation and
potential speed gains due to vectorization.  However, we have organized the
algorithms of \code{CorrelationFunctions.jl} to be able to utilize any patterns.
This was achieved by organizing computational functions not around a shifting
pattern, but rather around precomputed points representing triangle's vertices
-- thus making our approach under the hood using the ideas behind the weblike
pattern method.

@@

\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.8\linewidth]{images/web-like.png}
  \caption[]{Schematic representation of patterns for 3-point correlation
    functions evaluation: arbitrary or random triangle, weblike, and
    right-triangular patterns. Note that real patterns are more complex and are
    made simple for visualization, i.e., for right-triangular pattern we sample
    all possible variations between points. Equilateral or isosceles triangles
    can be considered as arbitrary patterns and can be either fixed in space, or
    rotated randomly.}
  \label{fig:pattern}
\end{figure*}

@@

\subsection{Comparison of sampling patterns}
\label{sec:compare_patterns}
To compare different sampling and averaging patterns we performed computations
of $S_3$ on Sandstone 3D sample.  They were performed in three different
manners: 1) isosceles triangle with a right-triangle between the equal side --
this sampling resembles both our right-triangle pattern and pattern of Malmir et
al. \cite{malmir2018}, computations were made within XY planes within the image
and, thus, represent directional CF; 2) equilateral triangle with side $r$
directionally analyzed within XY planes; 3) equilateral triangle with side $r$
averaged over 20 random rotations -- reminiscence of weblike and random
patterns.

@@

\subsection{Applications to material images}
\begin{table*}[!htp]
  \centering
  \begin{tabular}{|c|c|c|c|c|}
    \hline
    Sample name & Image type & Dimensions (pixels/voxels) & Image resolution ($\mu m$)
    & $Q_{0.5}$ \\
    \hline
    Sandstone & XCT &  $500 \times 500 \times 500$ & 2.25 & 0.9703 \\
    Soil & Reconstruction & $500 \times 500 \times 500$ & 0.72 & 0.9431 \\
    Shale & Reconstruction & $160 \times 160 \times 160$ & 2.25 & 0.8912 \\
    Sandstone & SEM &  $1280 \times 869$ & 0.10 & 0.9613 \\
    \hline
  \end{tabular}
  \caption{Summary of samples used for calculation of correlation functions.}
  \label{tab:summary}
\end{table*}
In this section we compute correlation functions for 2- and 3-dimensional porous
media samples. Table \ref{tab:summary} contains information about these
samples. Note that we compute surface correlation functions only for those
samples for which the criterion $Q_{0.5} > 0.97$ holds (the two sandstones and
the soil) \cite{samarin2023robust}. Images of the samples are presented on
\cref{fig:samples}. On plots of correlation functions for isotropic samples we
provide a family of curves with the first argument of the correlation function
being fixed. For anisotropic samples we provide heatmaps of correlation
functions where anisotropy can be easily observed.
\begin{figure*}[tp]
  \centering
  \subfigure[Sanstone XCT]{
    \includegraphics[width=0.4\linewidth]{images/15581.png}
    \label{fig:sample-sandstone-xct}}
  \hfill
  \subfigure[Sandstone SEM (different clusters of void phase are shown in color)]{
    \includegraphics[width=0.4\linewidth, frame]{images/sandstone1.png}
    \label{fig:sample-sandstone}}
  \vskip\baselineskip
  \subfigure[Soil]{
    \includegraphics[width=0.4\linewidth]{images/soil.png}
    \label{fig:sample-soil}}
  \hfill
  \subfigure[Shale]{
    \includegraphics[width=0.4\linewidth]{images/shale.png}
    \label{fig:sample-shale}}
  \caption[]{Images of samples used for calculation of higher-order correlation
    functions.}
  \label{fig:samples}
\end{figure*}

@@

We have made execution time comparisons of our algorithm with existing
algorithms.  However, it is also beneficial to compare some algorithmic details
against existing digital approaches -- proposed by Malmir~et.~al.
\cite{malmir2018} and another one by Smith and Torquato \cite{SMITH1988176}. The
first algorithm generates a triangle with fixed sides $r_1$ and $r_2$ and an
arbitrary angle $\theta$ between them and places it in the medium firstly making
a random translation and a rotation around a random vector by a random angle. By
making a large number of trials the algorithm gets a value of a three-point
correlation function at the point $(r_1, r_2, \theta)$. The second algorithm
takes a precalculated triangle $(r_1, r_2, \theta)$ from a weblike pattern and
places it in the medium without rotation, only applying a random
translation. This way the algorithm preserves orientation of triangles stored in
the pattern. The main advantage of our approach that can operate on arbitrary patterns is in
its parallelizability and ease of implementation. The rotations necessary
for arbitrary placement of the pattern within the image can be random as in the
Malmir et al. algorithm, or it can be a single rotation to align the sampling
triangle with the direction of anisotropy (\cref{fig:aniso}). The latter allows
computing directional 3-point correlation functions for anisotropic structures
similar to 2-point counterparts \cite{10.1063/1.4867611,EPL1}.
